Dual host-parasite transcriptomes of apicomplexan Eimeria falciformis and its natural mouse host

### 2016-08-08 figuresANDmanuscript/manuscriptBMCformat/manuscript_rnaseq.tex
Manuscript changes while working on feature counts integration and simple csv output to be handed over between scripts

\author[
%	addressref={aff1},
%	noteref={n1},
]{\inits{TK}\fnm{Totta} \snm{Kasemo}}
\author[
%	addressref={aff1},
]{\inits{SS}\fnm{Simone} \snm{Spork}}
\author[
%	addressref={aff2},
%	noteref={n1},
]{\inits{CD}\fnm{Christoph} \snm{Dieterich}}
\author[
%	addressref={aff1},
]{\inits{RL}\fnm{Richard} \snm{Lucius}}
\author[
%	addressref={aff1, aff3},                   % id's of addresses, e.g. {aff1,aff2}
	corref={aff1, aff3},                       % id of corresponding address, if any
%	noteref={n1},                        % id's of article notes, if any
	email={emanuel.heitlinger@hu-berlin.de}   % email address
]{\inits{EH}\fnm{Emanuel} \snm{Heitlinger}}

@@

\address[id=aff1]{%                           % unique id
	\orgname{Department of Biology, Humboldt-Universitat zu Berlin}, % university, etc
	\street{Philippstr. 13, Haus 14},                     %
	\postcode{10115}                                % post or zip code
	\city{Berlin},                              % city
	\cny{Germany}                                    % country
}
\address[id=aff2]{%  
	\orgname{Max Planck Institute for Biology of Ageing},
	\street{Joseph-Stelzmann-Strasse 9B},
	\postcode{50931}
	\city{Cologne},
	\cny{Germany}
}
\address[id=aff3]{%  
	\orgname{Department Reproduction Biology and Evolutionary Ecology, Leibniz Institute of Zoo and Wildlife Research},
	\street{Alfred-Kowalke-Str. 17},
	\postcode{10315}
	\city{Berlin},
	\cny{Germany}
}

@@

  \section*{Experimental overview}
We performed mRNA sequencing of caecum tissue from mice infected with
the apicomplexan parasite \textit{E. falciformis}. For each replicate
sample we enriched caecum tissue from three infected mice for
epithelial cells. We extracted mRNA and constructed sequencing
libraries for different timepoints after infection and mouse strains
with assumed different immunocompetence.
%% we have to bring in Simones data first with two sentences and a
%% figure or write it like this
Additonally we obtained data for challenge infections of mice
immunized one month prior to the experiment. For the parasite oocysts
and sporozoites were included as ``environmental'' lifecycle stages.

@@

@@

On the other end of the spectrum mice not previously immunized
featured an mRNA ``holo-transcriptome'' at 7dpi with the majority of
mRNA originating from the parasite as opposed to the host (Figure 1).

% latex table generated in R 3.3.1 by xtable 1.8-2 package
% Fri Aug  5 13:00:01 2016
\begin{table}[ht]
\caption{Experimental design, sequencing output and mapping statistics.}
\centering
\hspace*{-1.5cm}\begin{tabular}{lllllllll}
  \hline
Sample & \bcell{Sequencing\\method} & batch & \bcell{read mappings\\Mouse} & \bcell{read mappings\\\textit{E. falciformis}} & \bcell{Percentage\\\textit{E. falciformis}} & dpi & infection  \\ 
\hline
  NMRI\_2ndInf\_0dpi\_rep1 & GAII & 2 & 285032128 & 326 & 0 & 0dpi & Challenge \\ 
  Rag\_0dpi\_rep1 & hiseq & 3 & 43414004 & 474 & 0.001 & 0dpi & None \\ 
  C57BL6\_0dpi\_rep1 & hiseq & 3 & 53877840 & 491 & 0.001 & 0dpi & None \\ 
  C57BL6\_0dpi\_rep2 & hiseq & 3 & 76753491 & 657 & 0.001 & 0dpi & None \\ 
  Rag\_0dpi\_rep2 & hiseq & 3 & 80702547 & 730 & 0.001 & 0dpi & None \\ 
  NMRI\_2ndInf\_0dpi\_rep2 & hiseq & 3 & 90681785 & 3738 & 0.004 & 0dpi & Challenge \\ 
  \rowcolor{LightCyan}
  NMRI\_2ndInf\_3dpi\_rep1 & hiseq & 3 & 91352242 & 5355 & 0.006 & 3dpi & Challenge \\ 
  \rowcolor{LightCyan}
  NMRI\_2ndInf\_5dpi\_rep2 & hiseq & 3 & 113957667 & 12692 & 0.011 & 5dpi & Challenge \\ 
  \rowcolor{LightRed}
  NMRI\_1stInf\_0dpi\_rep1 & GAII & 1 & 229165002 & 31459 & 0.014 & 0dpi & None \\ 
  NMRI\_2ndInf\_3dpi\_rep2 & hiseq & 3 & 106446839 & 34699 & 0.033 & 3dpi & Challenge \\ 
  NMRI\_2ndInf\_7dpi\_rep2 & hiseq & 3 & 122999109 & 46700 & 0.038 & 7dpi & Challenge \\ 
  Rag\_1stInf\_5dpi\_rep1 & hiseq & 3 & 71173382 & 73445 & 0.103 & 5dpi & First \\ 
  Rag\_1stInf\_5dpi\_rep2 & hiseq & 3 & 57192380 & 113673 & 0.198 & 5dpi & First \\ 
  C57BL6\_1stInf\_5dpi\_rep1 & hiseq & 3 & 65523435 & 217606 & 0.331 & 5dpi & First \\ 
  Rag\_2ndInf\_5dpi\_rep1 & hiseq & 3 & 85288323 & 224273 & 0.262 & 5dpi & Challenge \\ 
  C57BL6\_1stInf\_5dpi\_rep2 & hiseq & 3 & 32887009 & 250954 & 0.757 & 5dpi & First \\ 
  C57BL6\_2ndInf\_5dpi\_rep1 & hiseq & 3 & 62053975 & 311900 & 0.500 & 5dpi & Challenge \\ 
  NMRI\_1stInf\_3dpi\_rep1 & GAII & 1 & 209723287 & 815820 & 0.388 & 3dpi & First \\ 
  NMRI\_1stInf\_3dpi\_rep2 & GAII & 2 & 515516785 & 903330 & 0.175 & 3dpi & First \\ 
  NMRI\_1stInf\_5dpi\_rep2 & GAII & 2 & 316721928 & 1609009 & 0.505 & 5dpi & Firs \\ 
  NMRI\_2ndInf\_5dpi\_rep1 & GAII & 2 & 589643389 & 3752923 & 0.632 & 5dpi & Challenge \\ 
  NMRI\_2ndInf\_7dpi\_rep1 & hiseq & 3 & 97221189 & 9927803 & 9.265 & 7dpi & Challenge \\ 
  NMRI\_1stInf\_5dpi\_rep1 & GAII & 2 & 204647381 & 18549727 & 8.311 & 5dpi & First \\ 
  NMRI\_sporozoites\_rep2 & GAII & 0 & 8702 & 21508353 & 99.960 & \textit{in vitro} & \textit{in vitro} \\ 
  NMRI\_1stInf\_5dpi\_rep3 & GAII & 0 & 334671421 & 54022504 & 13.899 & 5dpi & First \\ 
  NMRI\_1stInf\_7dpi\_rep1 & GAII & 1 & 12532238 & 79648900 & 86.405 & 7dpi & First \\ 
  NMRI\_sporozoites\_rep1 & GAII & 1 & 13800 & 92259539 & 99.985 & \textit{in vitro} & \textit{in vitro} \\ 
  NMRI\_oocysts\_rep1 & GAII & 1 & 11676 & 108477484 & 99.989 & envir. & envir. \\ 
  NMRI\_oocysts\_rep2 & GAII & 0 & 19024 & 126543533 & 99.985 & envir. & envir. \\ 
  NMRI\_1stInf\_7dpi\_rep2 & GAII & 1 & 94310278 & 154343046 & 62.072 & 7dpi & First \\ 
   \hline
\hspace*{+1.5cm}\end{tabular}
\begin{tablenotes}
  \small
\item *envir. ``environmental'' oocysts sporulated in potassium dichromate. \\
\item *\textit{in vitro} in vitro sporulated oocysts
\end{tablenotes}
\end{table}

@@

%% Before was not pretty. All those NA's. If people are like me they
%% hate NAs in tables ;-). I think you sould make the above table from
%% data with xtable.

%% best use a straight 2d table with...

% label & mouse strain & parasite dpi & naive/challenge & number of
% reads & #mapped to mouse & #mapped to \textit{E. falciformis}

%% column name "naive/challenge" can change

%% as technical as it seems people will want to know what our labels
%% mean (they appear in figures throughout) therfore it is good to
%% have them in table 1

@@

Statistical testing for differences between infected and uninfected
mice, revealed changes in mRNA abundance becoming more pronounced
(involving more genes) at later timepoints of infection (Table 1).

The observed fold-changes were in agreement with previously published
microarry data. Fold-Change data obtained from mice at 6dpi on Agilent
microarrays (Schmidt et al 2012) or 7dpi (our RNAseq data) shows a
strong correlateion (Spearman's \rho = 333; $R^2$ from a linear model
= 444 ; see also figure 2). Considering both and technical differences
between the tow methods and the biological changes along the lifecycle
this confirms the adequacy of our methods for assessing the host
transcriptome.

We selected all mouse genes differently expressed (FDR<1\%) between
uinfected and any of the infected samples.

%%% This is from methods but all the results should be presented here:

Those changes are mainly related to the onset of an
adaptive immune response.

\section*{The lifecycle of Eimeria is characterized by large changes in mRNA abundance}

%%% This is from methods but all the results should be presented here:

\clearpage
\section{mRNA abundance differences between different experimental groups}
\setlength{\tabcolsep}{8pt}
\begin{table}[H]
\small
\begin{center}
\caption{mRNA abundance differences between different experimental groups.}
%\label{tab:table3}
\begin{tabular}{*3l}    \toprule
\textit{Day post infection} & \textit{Ef} genes different & Mouse genes different \\ 
	\textit{comparisons} 	    & (FDR$\leq$1\%) &  (FDR$\leq$1\%/5\%) \\ \midrule
	NMRI 0 vs NMRI 3		& NA   & 274 \\
	NMRI 0 vs NMRI 5		& NA   & 1736 \\
	NMRI 0 vs NMRI 7		& NA   & 2802 \\
	NMRI 3 vs NMRI 5     		& 111  & 1 \\
	NMRI 3 vs NMRI 7  		& 1385 & 1407 \\ 
	NMRI 5 vs NMRI 7  		& 1895 & 873 \\ 
	C57BL/6 0 vs C57BL/6 5		& NA	& 914 \\
	Rag1-/- 0 vs Rag1-/- 5		& NA	& 45 \\ \midrule
\textit{Day post infection,} & 		 & 		 \\ 
\textit{parasite relevant comparisons} 	    & 		&  \\ \midrule
	Oocysts vs NMRI 3  	& 3310 & NA \\  
	Oocysts vs NMRI 5	& 3605 & NA \\ 
	Oocysts vs NMRI 7	& 3085 & NA \\ 
	Oocysts vs sporozoites  & 3421 & NA \\
	Sporozoites vs NMRI 3 	& 1663 & NA \\
	Sporozoites vs NMRI 5 	& 1605 & NA \\
	Sporozoites vs NMRI 7 	& 2473 & NA \\ \midrule
\textit{First and second infection} & 		 & 	 \\ 
\textit{comparisons} 	    & 		& 	\\ \midrule
	NMRI 3 1st vs NMRI 3 2nd  	& 0  & 5 \\
	NMRI 5 1st vs NMRI 5 2nd  	& 0  & 1 \\
	NMRI 7 1st vs NMRI 7 2nd  	& 0  & 902 \\
	C57BL/6 1st vs C57BL/6 2nd (day 5) & 0 &  mouse \\
	Rag1-/- 1st vs Rag1-/- 2nd (day 5) & 0 & mouse  \\ \midrule
\textit{Mouse strain comparisons} & 		 & 	 \\ \midrule
	NMRI vs C57BL/6 		& 22 	& NA* \\		
	NMRI vs Rag1-/-     		& 0 & NA* 	\\
	C57BL/6 vs Rag1-/-  		& 0   & 356 	\\ \midrule
\bottomrule
	\hline
\end{tabular}
\end{center}
\end{table}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage
\subsection{Life cycle analysis... \textit{E. falciformis}}
\begin{figure}[h!]
%\begin{center}
%\centering
	\includegraphics[width=\linewidth]{EfLifecycleHeatmap.pdf}  
	\caption{\textit{E. falciformis} mRNAs with significantly different abundance at different times p.i. in NMRI mice (see Table 2). Samples cluster according to expected developmental state of the parasite, without distinguishing between first and challenge infection. Hierarchical clustering was set to apply seven gene clusters.}
%\end{center}
\end{figure}

@@

\clearpage
\subsection{Parasite development reflected in mouse data}
\begin{figure}[h!]
%\begin{center}
%\centering
\includegraphics[width=\linewidth]{MmLifecycleHeatmap.pdf}  
\caption{Mouse mRNAs with significantly different abundance at different times p.i. in NMRI mice (see Table 2). Samples cluster according to expected developmental state of the parasite, without distinguishing between first and challenge infection. Rag1-/- mouse samples cluster together. Hierarchical clustering was set to apply four gene clusters.}
%\end{center}
\end{figure}

@@

\clearpage
\subsection*{GO term enrichments in heatmap clusters}
The annotations referred to here are inferred from orthologs in other \textit{Eimeria} spp. or
in \textit{T. gondii}

@@

A Thrombospondin type 1 domain-containing protein ortholog is highly abundant in cluster 4 (high 
abundance in oocysts). Thrombospondin type 1 domains have been reported in \textit{E. tenella} microneme
localizing proteins, MIC, e.g. MIC4 (Tomley01) In \textit{E. tenella} MIC4 mRNA was reported in 
sporozoites where it localizes to the apical end, and in late schizonts and late oocyst stages,
when sporozoites are forming. (Tomley01). For the same gene, the \textit{T. gondii} annotation is 
Sushi domain-containing protein, which is also the ortholog annotation of another gene in this 
cluster. In the related malaria parasite \textit{P. falciparum} the apical sushi 
protein, ASP, (which has a sushi domain) localizes to micronemes in merozoites but not 
other stages (OKeeffe05). 
Limulus clotting factor C, Coch-5b2 (Cochlin) and Lgl1, LCCL, 
(syn. F5/8 domain) domains are associated discoidin lectin domains and thereby with adhesion.
(Pfam entries for 'LCCL domain' and 'Discoidin domain', May 2016). 
Taken together, this indicates that the thrombospondin and sushi-domain genes 
(EfaB\_MINUS\_4114.g412 and EfaB\_PLUS\_1425.g183) are involved in sporozoite invasion in 
\textit{E. falciformis} and that the mRNAs are transcribed and available before excystation. 
A role in merozoite re-invasion in \textit{E. falciformis} is not indicated by our data.
The LCCL domain annotation and thrombospondins role in higher eukaryotes also indicates 
that adhesion or preparation for adhesion is important in oocysts. 
We suggest (speculate...?) that the thrombospondin annotated ortholog and the LCCL domain-containing protein 
(EfaB\_MINUS\_11233.g986) are involved in cell adhesion in \textit{E. falciformis}.
%One examples
%is found in the slime mold \textit{Dictyostelium discoideum)} in which the LCCL domain
%is part of the mold's discoidin adhesion protein. (Pfam entry for 'discoidin domain', May 2016)

@@

\subsection{Differentially abundant mRNAs used for hierarchical clustering}
All analyses were performed in R (cite R-core). Complete scripts are
available at git@github.com:derele/Ef\_RNAseq.git tagged as version
1.0. 
%% We will make this available as a git webpage probably

@@

%\begin{figure}
 %\begin{picture}
 % \includegraphics[width=0.5\linewidth]{ef-heatmap}
 %\end{picture}
%\end{figure}

\section*{Figures}
\begin{figure}[h!]
%  \includegraphics[width=0.5\linewidth]{efmm-ratio}
  \caption{\csentence{Parasite read sequences per mouse read sequences for each experimental condition.}
  The more advanced the infection is, the higher the ratio of parasite reads is, reflecting parasite replication between day zero and day seven. ......compare 1st and 2nd ....... patterns between Rag and NMRI/C57BL/6?  ........ Values are mean of replicates (n=2, if * n=1) on log10 scale. Each sample (replicate) consists of mRNA from three different mice.}
  \end{figure}

 % \begin{figure}[h!]
 % \includegraphics[width=0.7\linewidth]{ef-heatmap}
 % \caption{\csentence{Parasite genes with different mRNA abundance between samples, sorted into five clusters by hierarchical clustering (method = complete, distance = euclidean).}
 %     \textit{E. falciformis} samples from seven days p.i. cluster together (NMRI mice only). These samples have distinct mRNA abundance patterns in all gene clusters, although more pronounced in clusters 1 (up) and 4 (down). Distinct groups of genes also define sporozoites (cluster 5, up) and oocysts (cluster 2, up). mRNA profiles on days three and five p.i. from all three mouse strains cluster together. These samples are distinct from oocysts, NMRI day 7 p.i., and sporozoites, however closest to the latter. On scale bar, 0 is mean mRNA abundance for each gene. Up and downregulation is standard deviations from mean.}
 %     \end{figure}

%\begin{figure}[h!]
%  \includegraphics[width=0.7\linewidth]{mm-heatmap}
%  \caption{\csentence{Mouse genes with different mRNA abundance between samples, sorted into four clusters by hierarchical clustering (method = complete, distance = euclidean).}
%Three out of four samples from day 7 p.i. (NMRI only) cluster together. These samples are characterized by genes in cluster 2 (up), 3, and 4 (down). NMRI and C57BL/6 uninfected samples cluster together (left), defined by clusters 3, 4 (up), and 1 (down). Rag1-/- samples cluster together. These samples share a weak downregulation of most genes in cluster 2 as well as upregulation of a small group genes in the same cluster. Uninfected Rag1-/- samples are separated from infected ones with distinct profiles in clusters 1, 3, and 4. On scale bar, 0 is mean mRNA abundance for each gene. Up and downregulation is standard deviations from mean.}
%      \end{figure}

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
\begin{table}[h!]
\caption{Sample table title. This is where the description of the table should go.}
      \begin{tabular}{cccc}
        \hline
           & B1  &B2   & B3\\ \hline
        A1 & 0.1 & 0.2 & 0.3\\
        A2 & ... & ..  & .\\
        A3 & ..  & .   & .\\ \hline
      \end{tabular}
\end{table}

>>>>

\author[ <addressref>={aff1},] {\inits{TK}\fnm{Totta} \snm{Kasemo}}
\author[ <addressref>={aff1},] {\inits{SS}\fnm{Simone} \snm{Spork}}
\author[ <addressref>={aff2},] {\inits{CD}\fnm{Christoph} \snm{Dieterich}}
\author[ <addressref>={aff1},]{\inits{RL}\fnm{Richard} \snm{Lucius}}
\author[ <addressref>={aff1, aff3},              
  corref={aff1},                   
  email={emanuel.heitlinger@hu-berlin.de}]{\inits{EH}\fnm{Emanuel} \snm{Heitlinger}}

@@

\address[id=aff1]{
	\orgname{Institute of Biology, Humboldt-Universitat zu Berlin}, 
	\street{Philippstr. 13, Haus 14},
	\postcode{10115}                 
	\city{Berlin},                   
	\cny{Germany}                    
}
\address[id=aff2]{  
	\orgname{Max Planck Institute for Biology of Ageing},
	\street{Joseph-Stelzmann-Strasse 9B},
	\postcode{50931}
	\city{Cologne},
	\cny{Germany}
}
\address[id=aff3]{  
	\orgname{Leibniz Institute of Zoo and Wildlife Research},
	\street{Alfred-Kowalke-Str. 17},
	\postcode{10315}
	\city{Berlin},
	\cny{Germany}
}

@@

  \section*{A dual transcriptomics experiment}
We performed mRNA sequencing of caecum tissue from mice infected with
the apicomplexan parasite \textit{E. falciformis}. We worked with two
biological replicates for all but two conditions (with one and three
replicates, respecitively). For each replicate sample we enriched
caecum tissue from three infected mice for epithelial cells. We
extracted mRNA and constructed sequencing libraries for different
timepoints after infection and mouse strains which displayed different
immunocompetence in infection trials (see figure x).

Briefly RAG-1-/- knockout mice, in which mature B and T lymphocytes
are absent showed ...

Additonally we obtained data for challenge infections of mice
immunized one month prior to the experiment.

For the parasite oocysts and sporozoites were included as
``environmental'' and ``\textit{in vitro}'' processed lifecycle
stages.

@@

The results of these multivariate analyses in relation to biological
interpretation are described further below.

@@

One sample (NMRI\_1stInf\_0dpi\_rep1) was additonally excluded because
the uninfected control showed an unexpected mapping of reads to the
\textit{E. falciformis} genome. We consider this together with the
unsuccessfully infected samples above to display an undecided state of
infection.

On the other end of the spectrum mice not previously immunized
featured an mRNA ``holo-transcriptome'' at 7dpi with the majority of
mRNA originating from the parasite as opposed to the host (Figure
1). The overall mRNA output of the sampled caecum tissue at this point
of infection is dominated by the parasite with proportional mRNA
abundance between XXX and XXX %

@@

@@

Statistical testing for differetial expression between infected and
uninfected mice, revealed changes in mRNA abundance becoming more
pronounced (involving more genes) at later post infection (especially
at 7dpi; Table 1).

The observed changes between controls and 7dpi were in agreement with
previously published microarry data. Fold-Change data obtained from
mice at 6dpi on Agilent microarrays (Schmidt et al 2012) or 7dpi (our
RNAseq data) agains uninfected controls shows a strong correlateion
(Spearman's \rho = 333; $R^2$ from a linear model = 444 ; figure 2
a). Considering both and technical differences between the tow methods
and the biological changes along the lifecycle this confirms the
adequacy of our methods for assessing the host transcriptome.

XX mouse genes were considered differently expressed (DE; FDR<1\%)
between uinfected and 3dpi, XX genes between controls and 5dpi, xxx
genes between 0dpi and 7dpi. This lead to combined set of xxxx genes
reacting to infection (see figure 2 b).

We detected no distinct response early in infection.  When testing for
differences between 3dpi and later infection (5dpi and 7dpi) DE genes
were simply a subset of genes DE between respective infections and
controls.

To further validate these patterns we performed hierarchical
clustering on mouse mRNAs DE at different dpi (union of DE
genes).

%% Samples from day 7 p.i. form a separate cluster (NMRI mice only), not
%% distinguishing between first and challenge infection. For these
%% samples, the pattern is most pronounced in gene cluster 1 (up), and 4
%% and 6 (down). Within cluster 1, the second replicate from the
%% challenge infection has a deviant profile (compare mouse data in
%% Figure 2). Distinct groups of genes also define sporozoites (cluster 5
%% up) and oocysts (cluster 3 up, cluster 6 down). mRNA profiles on days
%% 3 and 5 p.i. from all three mouse strains cluster together. These
%% samples are distinct from oocysts, NMRI day 7 p.i., and
%% sporozoites. Sporozoites cluster closer with days 3 and 5 than with
%% oocysts or day 7.

%%% Test _this_:
The more ponounced changes later in infection are is likely reflecting
a lag time in detection before the parasite reaches sufficient
intensity during asexual expansion (schizogony). Those changes are
mainly related to the onset of an adaptive immune response.

\section*{The lifecycle of Eimeria is characterized by major changes in mRNA abundance}

@@

@@

@@

\subsection*{GO term enrichments in heatmap clusters}
The annotations referred to here are inferred from orthologs in other \textit{Eimeria} spp. or
in \textit{T. gondii}

@@

A Thrombospondin type 1 domain-containing protein ortholog is highly
abundant in cluster 4 (high abundance in oocysts). Thrombospondin type
1 domains have been reported in \textit{E. tenella} microneme
localizing proteins, MIC, e.g. MIC4 (Tomley01) In \textit{E. tenella}
MIC4 mRNA was reported in sporozoites where it localizes to the apical
end, and in late schizonts and late oocyst stages, when sporozoites
are forming. (Tomley01). For the same gene, the \textit{T. gondii}
annotation is Sushi domain-containing protein, which is also the
ortholog annotation of another gene in this cluster. In the related
malaria parasite \textit{P. falciparum} the apical sushi protein, ASP,
(which has a sushi domain) localizes to micronemes in merozoites but
not other stages (OKeeffe05).  Limulus clotting factor C, Coch-5b2
(Cochlin) and Lgl1, LCCL, (syn. F5/8 domain) domains are associated
discoidin lectin domains and thereby with adhesion.  (Pfam entries for
'LCCL domain' and 'Discoidin domain', May 2016).  Taken together, this
indicates that the thrombospondin and sushi-domain genes
(EfaB\_MINUS\_4114.g412 and EfaB\_PLUS\_1425.g183) are involved in
sporozoite invasion in \textit{E. falciformis} and that the mRNAs are
transcribed and available before excystation.  A role in merozoite
re-invasion in \textit{E. falciformis} is not indicated by our data.
The LCCL domain annotation and thrombospondins role in higher
eukaryotes also indicates that adhesion or preparation for adhesion is
important in oocysts.  We suggest (speculate...?) that the
thrombospondin annotated ortholog and the LCCL domain-containing
protein (EfaB\_MINUS\_11233.g986) are involved in cell adhesion in
\textit{E. falciformis}.
%One examples is found in the slime mold \textit{Dictyostelium
%discoideum)} in which the LCCL domain is part of the mold's discoidin
%adhesion protein. (Pfam entry for 'discoidin domain', May 2016)

@@

\subsection{Differentially abundant mRNAs used for hierarchical clustering}
All analyses were performed in R (cite R-core). Complete scripts are
available at https://github.com/derele/Ef\_RNAseq.git tagged as version
1.0.
%% We will make this available as a github webpage!!!

@@

\section*{Figures}

\begin{figure}[h!]
%  \includegraphics[width=0.5\linewidth]{efmm-ratio}
  \caption{\csentence{Experimental design and outcome of a dual RNAseq experiment}
    read sequences per mouse read sequences for each experimental
    condition. The more advanced the infection is, the higher the
    ratio of parasite reads is, reflecting parasite replication
    between day zero and day seven. ......compare 1st and 2nd
    ....... patterns between Rag and NMRI/C57BL/6?  ........ Values
    are mean of replicates (n=2, if * n=1) on log10 scale. Each sample
    (replicate) consists of mRNA from three different mice.}
  \end{figure}

\subsection{Life cycle analysis... \textit{E. falciformis}}
\begin{figure}[h!]
%\begin{center}
%\centering
	\includegraphics[width=\linewidth]{EfLifecycleHeatmap.pdf}  
	\caption{\textit{E. falciformis} mRNAs with significantly
          different abundance at different times p.i. in NMRI mice
          (see Table 2). Samples cluster according to expected
          developmental state of the parasite, without distinguishing
          between first and challenge infection. Hierarchical
          clustering was set to apply seven gene clusters.}
%\end{center}
\end{figure}

\subsection{Parasite development reflected in mouse data}
\begin{figure}[h!]
%\begin{center}
%\centering
\includegraphics[width=\linewidth]{MmLifecycleHeatmap.pdf}  
\caption{Mouse mRNAs with significantly different abundance at
  different times p.i. in NMRI mice (see Table 2). Samples cluster
  according to expected developmental state of the parasite, without
  distinguishing between first and challenge infection. Rag1-/- mouse
  samples cluster together. Hierarchical clustering was set to apply
  four gene clusters.}
%\end{center}
\end{figure}

% \begin{figure}[h!]
 % \includegraphics[width=0.7\linewidth]{ef-heatmap} \caption{\csentence{Parasite
 % genes with different mRNA abundance between samples, sorted into
 % five clusters by hierarchical clustering (method = complete,
 % distance = euclidean).}  \textit{E. falciformis} samples from seven
 % days p.i. cluster together (NMRI mice only). These samples have
 % distinct mRNA abundance patterns in all gene clusters, although
 % more pronounced in clusters 1 (up) and 4 (down). Distinct groups of
 % genes also define sporozoites (cluster 5, up) and oocysts (cluster
 % 2, up). mRNA profiles on days three and five p.i. from all three
 % mouse strains cluster together. These samples are distinct from
 % oocysts, NMRI day 7 p.i., and sporozoites, however closest to the
 % latter. On scale bar, 0 is mean mRNA abundance for each gene. Up
 % and downregulation is standard deviations from mean.}  \end{figure}

%\begin{figure}[h!]
%  \includegraphics[width=0.7\linewidth]{mm-heatmap} \caption{\csentence{Mouse
%  genes with different mRNA abundance between samples, sorted into
%  four clusters by hierarchical clustering (method = complete,
%  distance = euclidean).}  Three out of four samples from day 7
%  p.i. (NMRI only) cluster together. These samples are characterized
%  by genes in cluster 2 (up), 3, and 4 (down). NMRI and C57BL/6
%  uninfected samples cluster together (left), defined by clusters 3,
%  4 (up), and 1 (down). Rag1-/- samples cluster together. These
%  samples share a weak downregulation of most genes in cluster 2 as
%  well as upregulation of a small group genes in the same
%  cluster. Uninfected Rag1-/- samples are separated from infected
%  ones with distinct profiles in clusters 1, 3, and 4. On scale bar,
%  0 is mean mRNA abundance for each gene. Up and downregulation is
%  standard deviations from mean.}  \end{figure}

@@

%% Use of \listoftables is discouraged.
%%

\begin{table}[ht]
\caption{Experimental design, sequencing output and mapping statistics.}
\centering
\hspace*{-1.5cm}\begin{tabular}{lllllllll}
  \hline
Sample & \bcell{Sequencing\\method} & batch & \bcell{read mappings\\Mouse} & \bcell{read mappings\\\textit{E. falciformis}} & \bcell{Percentage\\\textit{E. falciformis}} & dpi & infection  \\ 
\hline
  NMRI\_2ndInf\_0dpi\_rep1 & GAII & 2 & 285032128 & 326 & 0 & 0dpi & Challenge \\ 
  Rag\_0dpi\_rep1 & hiseq & 3 & 43414004 & 474 & 0.001 & 0dpi & None \\ 
  C57BL6\_0dpi\_rep1 & hiseq & 3 & 53877840 & 491 & 0.001 & 0dpi & None \\ 
  C57BL6\_0dpi\_rep2 & hiseq & 3 & 76753491 & 657 & 0.001 & 0dpi & None \\ 
  Rag\_0dpi\_rep2 & hiseq & 3 & 80702547 & 730 & 0.001 & 0dpi & None \\ 
  NMRI\_2ndInf\_0dpi\_rep2 & hiseq & 3 & 90681785 & 3738 & 0.004 & 0dpi & Challenge \\ 
  \rowcolor{LightCyan}
  NMRI\_2ndInf\_3dpi\_rep1 & hiseq & 3 & 91352242 & 5355 & 0.006 & 3dpi & Challenge \\ 
  \rowcolor{LightCyan}
  NMRI\_2ndInf\_5dpi\_rep2 & hiseq & 3 & 113957667 & 12692 & 0.011 & 5dpi & Challenge \\ 
  \rowcolor{LightRed}
  NMRI\_1stInf\_0dpi\_rep1 & GAII & 1 & 229165002 & 31459 & 0.014 & 0dpi & None \\ 
  NMRI\_2ndInf\_3dpi\_rep2 & hiseq & 3 & 106446839 & 34699 & 0.033 & 3dpi & Challenge \\ 
  NMRI\_2ndInf\_7dpi\_rep2 & hiseq & 3 & 122999109 & 46700 & 0.038 & 7dpi & Challenge \\ 
  Rag\_1stInf\_5dpi\_rep1 & hiseq & 3 & 71173382 & 73445 & 0.103 & 5dpi & First \\ 
  Rag\_1stInf\_5dpi\_rep2 & hiseq & 3 & 57192380 & 113673 & 0.198 & 5dpi & First \\ 
  C57BL6\_1stInf\_5dpi\_rep1 & hiseq & 3 & 65523435 & 217606 & 0.331 & 5dpi & First \\ 
  Rag\_2ndInf\_5dpi\_rep1 & hiseq & 3 & 85288323 & 224273 & 0.262 & 5dpi & Challenge \\ 
  C57BL6\_1stInf\_5dpi\_rep2 & hiseq & 3 & 32887009 & 250954 & 0.757 & 5dpi & First \\ 
  C57BL6\_2ndInf\_5dpi\_rep1 & hiseq & 3 & 62053975 & 311900 & 0.500 & 5dpi & Challenge \\ 
  NMRI\_1stInf\_3dpi\_rep1 & GAII & 1 & 209723287 & 815820 & 0.388 & 3dpi & First \\ 
  NMRI\_1stInf\_3dpi\_rep2 & GAII & 2 & 515516785 & 903330 & 0.175 & 3dpi & First \\ 
  NMRI\_1stInf\_5dpi\_rep2 & GAII & 2 & 316721928 & 1609009 & 0.505 & 5dpi & Firs \\ 
  NMRI\_2ndInf\_5dpi\_rep1 & GAII & 2 & 589643389 & 3752923 & 0.632 & 5dpi & Challenge \\ 
  NMRI\_2ndInf\_7dpi\_rep1 & hiseq & 3 & 97221189 & 9927803 & 9.265 & 7dpi & Challenge \\ 
  NMRI\_1stInf\_5dpi\_rep1 & GAII & 2 & 204647381 & 18549727 & 8.311 & 5dpi & First \\ 
  NMRI\_sporozoites\_rep2 & GAII & 0 & 8702 & 21508353 & 99.960 & \textit{in vitro} & \textit{in vitro} \\ 
  NMRI\_1stInf\_5dpi\_rep3 & GAII & 0 & 334671421 & 54022504 & 13.899 & 5dpi & First \\ 
  NMRI\_1stInf\_7dpi\_rep1 & GAII & 1 & 12532238 & 79648900 & 86.405 & 7dpi & First \\ 
  NMRI\_sporozoites\_rep1 & GAII & 1 & 13800 & 92259539 & 99.985 & \textit{in vitro} & \textit{in vitro} \\ 
  NMRI\_oocysts\_rep1 & GAII & 1 & 11676 & 108477484 & 99.989 & envir. & envir. \\ 
  NMRI\_oocysts\_rep2 & GAII & 0 & 19024 & 126543533 & 99.985 & envir. & envir. \\ 
  NMRI\_1stInf\_7dpi\_rep2 & GAII & 1 & 94310278 & 154343046 & 62.072 & 7dpi & First \\ 
   \hline
\hspace*{+1.5cm}\end{tabular}
\begin{tablenotes}
  \small
\item *envir. ``environmental'' oocysts sporulated in potassium dichromate. \\
\item *\textit{in vitro} in vitro sporulated oocysts
\end{tablenotes}
\end{table}
